Manuscript for a Data Analysis Project: Diabetes Diagnosis

### 2024-06-22 code/eda-code/EDA-Code.qmd
Updated Statistical Analysis Code - Final Project Part 3
Worked on the statistical analysis code for part 3 of the final project.

Updated manuscript with new relevant information from statistical analysis process.

Updated EDA code to create histograms all in one image instead of seperate images.

```{r}
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(readr)
library(tidyr) # data wrangling
library(rpart) # DT
library(tidyverse)
library(car)
library(tree)
library(caret)
library(MASS)
library(ROCR)
library(randomForest)
library(corrplot)
library(ggcorrplot) # corr plot
library(reshape2) # heat map
library(GGally) # Pais ggplot2
```

@@

```{r}

```

@@

```{r}
p1 <- mydata %>% ggplot(aes(x=Diabetes_012)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures","diabetes-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

```{r}
p2 <- mydata %>% ggplot(aes(x=Smoker)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures","smoker-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

```{r}
p3 <- mydata %>% ggplot(aes(x=BMI)) + geom_histogram() 
plot(p3)
figure_file = here("results", "figures","BMI-distribution.png")
ggsave(filename = figure_file, plot=p3) 
```

```{r}
p4 <- mydata %>% ggplot(aes(x=PhysHlth)) + geom_histogram() 
plot(p4)
figure_file = here("results", "figures","PhysHlth-distribution.png")
ggsave(filename = figure_file, plot=p4) 
```

```{r}
p5 <- mydata %>% ggplot(aes(x=Age)) + geom_histogram() 
plot(p5)
figure_file = here("results", "figures","Age-distribution.png")
ggsave(filename = figure_file, plot=p5) 
```

>>>>

```{r}
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(readr)
library(tidyr) # data wrangling
library(rpart) # DT
library(tidyverse)
library(car)
library(tree)
library(caret)
library(MASS)
library(ROCR)
library(randomForest)
library(corrplot)
library(ggcorrplot) # corr plot
library(reshape2) # heat map
library(GGally) # Pais ggplot2
library(gridExtra)
```

@@

@@

```{r}
# Function to create histograms for each variable
create_histograms <- function(data) {
  plots <- list()
  for (col_name in names(data)) {
    if (is.numeric(data[[col_name]])) {
      p <- ggplot(data, aes_string(x = col_name)) + 
        geom_histogram(binwidth = 1, fill = "salmon", color = "black") +
        labs(title = col_name) +
        theme_minimal()
    } else {
      p <- ggplot(data, aes_string(x = col_name)) + 
        geom_bar(fill = "salmon", color = "black") +
        labs(title = col_name) +
        theme_minimal()
    }
    plots[[col_name]] <- p
  }
  return(plots)
}

# Create histograms for all variables
histograms <- create_histograms(mydata)

# Arrange the plots in a grid
n <- length(histograms)
nCol <- ceiling(sqrt(n))
nRow <- ceiling(n / nCol)
combined_plot <- marrangeGrob(grobs = histograms, nrow = nRow, ncol = nCol)

# Save the combined plot to a PNG file
figure_file <- here("results", "figures", "variable-distributions.png")
ggsave(filename = figure_file, plot = combined_plot, width = 20, height = 20)
```

### 2024-06-22 products/manuscript/Manuscript.qmd
Updated Statistical Analysis Code - Final Project Part 3
Worked on the statistical analysis code for part 3 of the final project.

Updated manuscript with new relevant information from statistical analysis process.

Updated EDA code to create histograms all in one image instead of seperate images.

|------------|------------------------------------------------------------|

@@

*\*To be added in later portion of project*

## Statistical analysis

For my analysis, I will first look at the distributions of each variable to identify if the dataset is balanced and outliers. I do not think any data cleaning will need to be done as that was completed when the data was collected from the CDCs website. Next, I will create a correlation matrix to view the correlations between each predictor and the response variable. After viewing the correlations, I will then use simple linear regression and random forest models to identify variables of importance, these variables will be noted and used later during the modeling phase. I will then split the data set into a training and test set using a 80/20 split and begin the modeling phase. This section will be filled out more once we are further into the class and have discussed more approaches/methods.

***\* Any section past results has not been modified yet. I only completed the sections required for the first part.***

@@

### Diabetes Distribution

```{r}
knitr::include_graphics(here("results","figures","diabetes-distribution.png"))
```

### BMI Distribution

```{r}
knitr::include_graphics(here("results","figures","BMI-distribution.png"))
```

### PhysHlth Distribution

```{r}
knitr::include_graphics(here("results","figures","PhysHlth-distribution.png"))
```

### Age Distribution

```{r}
knitr::include_graphics(here("results","figures","Age-distribution.png"))
```

### Smoker Distribution

```{r}
knitr::include_graphics(here("results","figures","smoker-distribution.png"))
```

@@

***\*To be added in later portion of project***

@@

***\*To be added in later portion of project***

@@

***\*To be added in later portion of project***

>>>>

|------------------|------------------------------------------------------|

@@

Not much occurs during the cleaning stage as the dataset was processed and cleaned prior. The only modifications made to the dataset was removing the 'ID' column as it is not useful for this analysis. Once the data was imported, it was only a matter of converting categorical variables to factors so they would function properly. One-hot encoding was implemented, but only for the multinomial logistic regression.

## Statistical analysis method

For my analysis, I will first look at the distributions of each variable to identify if the dataset is balanced and outliers. I do not think any data cleaning will need to be done as that was completed when the data was collected from the CDCs website. Next, I will create a correlation matrix to view the correlations between each predictor and the response variable. After viewing the correlations, I will then use simple linear regression and random forest models to identify variables of importance, these variables will be noted and used later during the modeling phase. I will utilize cross validation during the modeling phase. This section will be filled out more once we are further into the class and have discussed more approaches/methods.

@@

### Distribution of all variables

```{r}
knitr::include_graphics(here("results","figures","variable-distributions.png"))
```

@@

***\*IN PROGRESS***

### Methods

After processing and exploring the data, it was discovered that a majority of the features in this dataset are categorical. Therefore, we will be making use of models that are more rigorous to our situation, such as random forest and KNN. However before trying those models, we will first do a simple multinomial logistic regression. Why multinomial? That is because our response variable is not binary, it has three levels to it; typically, logistic regression is used for binary classification. After creating models with all features included, we will go on to look at models with features that fit our importance criteria (i.e. look at the models results when using the most important features).

### Preparing the data

For some of the models, mainly multinomial logistic regression, we need to further process our data. The method we will use is known as one-hot encoding. One-hot encoding is a process of converting categorical variables into a binary (0 or 1) format where each category becomes a new binary column. We need to do this because most machine learning algorithms require numerical input and cannot handle categorical data directly, so formatting this data in a different manner will provide us with better results.

### Results

#### Multinomial Logistic Regression - All Features

```         
Confusion Matrix and Statistics

          Reference
Prediction    X0    X1    X2
        X0 21075   398  2891
        X1     3     1     1
        X2   477    32   622

Overall Statistics

               Accuracy : 0.8509          
                 95% CI : (0.8465, 0.8553)
    No Information Rate : 0.8453          
    P-Value [Acc > NIR] : 0.006606

                  Kappa : 0.1995

 Mcnemar's Test P-Value : < 2.2e-16

Statistics by Class:

                     Class: X0 Class: X1 Class: X2
Sensitivity             0.9777 2.320e-03   0.17701
Specificity             0.1663 9.998e-01   0.97685
Pos Pred Value          0.8650 2.000e-01   0.54996
Neg Pred Value          0.5775 9.831e-01   0.88132
Prevalence              0.8453 1.690e-02   0.13780
Detection Rate          0.8265 3.922e-05   0.02439
Detection Prevalence    0.9555 1.961e-04   0.04435
Balanced Accuracy       0.5720 5.011e-01   0.57693
```

\- Notes (Will be expanded and explained in future update):

-   Accuracy: 85.09%

-   Target variable may not be balanced

#### Random Forest - All Features

```         
Confusion Matrix and Statistics

          Reference
Prediction    X0    X1    X2
        X0 21555   431  3514
        X1     0     0     0
        X2     0     0     0

Overall Statistics

               Accuracy : 0.8453          
                 95% CI : (0.8408, 0.8497)
    No Information Rate : 0.8453          
    P-Value [Acc > NIR] : 0.5042

                  Kappa : 0

 Mcnemar's Test P-Value : NA

Statistics by Class:

                     Class: X0 Class: X1 Class: X2
Sensitivity             1.0000    0.0000    0.0000
Specificity             0.0000    1.0000    1.0000
Pos Pred Value          0.8453       NaN       NaN
Neg Pred Value             NaN    0.9831    0.8622
Prevalence              0.8453    0.0169    0.1378
Detection Rate          0.8453    0.0000    0.0000
Detection Prevalence    1.0000    0.0000    0.0000
Balanced Accuracy       0.5000    0.5000    0.5000
```

\- Notes (Will be expanded and explained in future update):

-   Model is only predicting '0', likely an error in model setup

-   Due to inaccurate predictions, accuracy is not applicable.

#### KNN - All Features

-   in progress

@@

***\*To be updated in later portion of project***

@@

***\*To be updated in later portion of project***

-   LIMIT: Having a lot of categorical features

-   LIMIT: Having a response variable with 3 levels instead of a binary.
